---
title: "Data_Vis - 1/9/25"
output: github_document
date: "2025-01-09"
---

```{r}
library(palmerpenguins)
library(tidyverse)
```
# how many rows are in penguins?
```{r}
nrow(penguins)
ncol(penguins)
```
##What does bill_depth_mm mean?
```{r}
?penguins
```
#make a scatter plot of bill depth vs bill length and what is the relationship between them?
```{r}
ggplot(
   penguins,
   aes(x = bill_length_mm, y = bill_depth_mm, color = species)
   ) +
   geom_point() +
   geom_smooth(method = 'lm', se = F)
```

#scatter plot bill depth and species
```{r}
ggplot(
   penguins,
   aes(x = bill_depth_mm, y = species, color = species)
   ) +
   geom_boxplot()
```

#troubleshooting problem
```{r}
ggplot(data = penguins, aes(x=bill_length_mm, y=bill_depth_mm)) + 
  geom_point(outl.rm = T)
```

#what does na.rm do? It removes your na values; by default it normally does not remove them

#using labs() function -- labels things
```{r}
ggplot(data = penguins, aes(x=bill_length_mm, y=bill_depth_mm)) + 
  geom_point() + 
  labs(
    title = "Data from palmerpenguins",
    subtitle = "Penguins are dope"
  )
```

#Recreating a plot
```{r}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g)) + 
  geom_point(aes(color = bill_depth_mm)) +
  geom_smooth() 
```

#recreting plot with different color scheme
```{r}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g, colour = bill_depth_mm)) + 
  geom_point() +
  geom_smooth() 
```

#run this block of code and predict what it does
```{r}
ggplot(
      data = penguins,
      mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
    ) +
      geom_point() +
      geom_smooth(se = FALSE) +
      labs(
        title = "This graph is ugly")
```
