---
title: "Data_Vis - 1/9/25"
output: github_document
date: "2025-01-09"
editor_options: 
  chunk_output_type: console
---

```{r}
library(palmerpenguins)
library(tidyverse)
```
# how many rows are in penguins?
```{r}
nrow(penguins)
ncol(penguins)
```
##What does bill_depth_mm mean?
```{r}
?penguins
```
#make a scatter plot of bill depth vs bill length and what is the relationship between them?
```{r}
ggplot(
   penguins,
   aes(x = bill_length_mm, y = bill_depth_mm, color = species)
   ) +
   geom_point() +
   geom_smooth(method = 'lm', se = F)
```

#scatter plot bill depth and species
```{r}
ggplot(
   penguins,
   aes(x = bill_depth_mm, y = species, color = species)
   ) +
   geom_boxplot()
```

#troubleshooting problem
```{r}
ggplot(data = penguins, aes(x=bill_length_mm, y=bill_depth_mm)) + 
  geom_point(outl.rm = T)
```

#what does na.rm do? It removes your na values; by default it normally does not remove them

#using labs() function -- labels things
```{r}
ggplot(data = penguins, aes(x=bill_length_mm, y=bill_depth_mm)) + 
  geom_point() + 
  labs(
    title = "Data from palmerpenguins",
    subtitle = "Penguins are dope"
  )
```

#Recreating a plot
```{r}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g)) + 
  geom_point(aes(color = bill_depth_mm)) +
  geom_smooth() 
```

#recreting plot with different color scheme
```{r}
ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g, colour = bill_depth_mm)) + 
  geom_point() +
  geom_smooth() 
```

#run this block of code and predict what it does
```{r}
ggplot(
      data = penguins,
      mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
    ) +
      geom_point() +
      geom_smooth(se = FALSE) +
      labs(
        title = "This graph is ugly")
```

#Will these two graphs look different? Why/why not? - both make the same looking graph
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()
```

#Second peice of code
```{r}
ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```

#Why are the points not blue? -- anything put in aes(), treated as argument therefore blue is treated as a category. Need to put color = 'blue' in geom_point() or geom_smooth() which will either make the points blue or the treadline blue
```{r}
ggplot(
    data = penguins, 
    mapping = aes(x = bill_length_mm, y = bill_depth_mm, color = 'blue')
) +
    geom_point() + geom_smooth(method = "lm")

#Refined code
ggplot(
    data = penguins, 
    mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) +
    geom_point() + geom_smooth(method = "lm", color = 'blue')
```

#How do you bump up the point size? And increase transparency of the points?
```{r}
ggplot(
    data = penguins, 
    mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) +
    geom_point(
      size = 2,
      alpha = 0.5, #alpha changes the transparency of points
    ) + geom_smooth(method = "lm", color = 'blue')
```

#Make a bar plot of species of penguins, where you assign species to the y aesthetic. How is this plot different? -- orientation is different
```{r}
ggplot(
  penguins,
  aes(y = fct_infreq(species))
) +
  geom_bar()
```

#How are the following two plots different? Which aesthetic, color or fill, is more useful for changing the color of bars? -- color just outlines when using bar graph, but when using fill() this changes the entire color of the bar graph
```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```

#What does the bins argument in geom_histogram() do? -- specifies the width of the bins
```{r}
ggplot(
  penguins,
  aes(x = body_mass_g)
) +
  geom_histogram(binwidth = 2)
library(diamonds)
```

#...


#1.5.5
```{r}
ggplot(
  mpg,
  aes(x = hwy, y = displ, line) 
) +
  geom_point() + geom_smooth(linewidth = 10, color = 'red')
```

#1.5.5 Q6
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() 
  #labs(color = "Species") # just need to comment out this line.
```

#1.5.5 Q7
```{r}
#first piece of code
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")

#second piece of code
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```